Planarity and Street Network Representation in Urban Form Analysis

### 2018-02-16 main.tex
update fig and author
rework quotes table into text

\usepackage{draftwatermark}
\SetWatermarkText{DRAFT}
\SetWatermarkScale{1}
\SetWatermarkLightness{0.9}

@@

\runninghead{Geoff Boeing}
\title{Planarity and Street Network Representation in Urban Form Analysis}
\author{Geoff Boeing \affilnum{1}}
\affiliation{\affilnum{1}Affiliation redacted}
\corrauth{Author Redacted, Address Redacted}
\email{Email Redacted}

@@

\keywords{street network, GIS, urban form, transportation, urban design}

@@

In urban planning and transportation research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity. Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.

@@

\section{Background and Motivation}

@@

\begin{table*}[htbp]
	\centering
	\caption{Survey of recent statements in the research literature regarding the representation of street networks as planar graphs.}
	\label{tab:planar_quotes}
	\input{figures_tables/table_planar_quotes.tex}
\end{table*}

\begin{figure}[tbp]
	\center
	\includegraphics[width=0.48\textwidth]{figures_tables/planar_vs_not.png}
	\caption{Formally and spatially planar (left) and nonplanar (right) road networks.}
	\label{fig:planar_vs_not}
\end{figure}

@@

In the urban studies literature, street networks are often modeled as planar graphs. Many street networks are obviously nonplanar in reality, but researchers build these planar models as a useful simplification of real-world complexity. Table \ref{tab:planar_quotes} surveys recent statements and reasoning around this approach. Some scholars argue that street networks are planar graphs, while others prefer to hedge slightly, arguing that street networks are \emph{approximately} or \emph{essentially} planar but close enough to be well-modeled as such.

@@

If a planar graph models a street network poorly, it could do so in multiple ways. Imposing planarity on a nonplanar street network produces artificial nodes in the graph at bridges and tunnels, which breaks routing. For example, this misrepresentation would allow a routing engine to recommend a left turn at an \enquote{intersection} that is actually an overpass \citep[p.~6]{kwan_review_1996}. Intersection counts and densities would be overestimated in the presence of these false nodes. Consequently, edge lengths would be underestimated due to these artificial breakpoints splitting up street segments. Finally, this bias might behave inconsistently across different kinds of cities and street network types based on the extents to which they are planar.

Given these issues, there are some unanswered questions in the empirical literature. What do \emph{approximately planar} and \emph{well-modeled} mean for street network research? How do planar simplifications impact the results of street network analyses? Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities (or neighborhoods) to different extents? And if street networks are at least generally approximately planar, how can we measure the extent to which a given street network is nonplanar?

The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,chimani_non-planar_2009}. However, these measures are imperfect and hard to compute \citep{szekely_successful_2004,chimani_vertex_2012}. They also fail to adjust for the size, density, or real-world embedding of a spatial network. In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks no widely-accepted measure of the extent of nonplanarity has emerged and calls for the development of better indicators.

@@

This study builds on this rich body of street network research by developing two new measures of the extent to which a spatial network is planar. It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results (focusing on the common urban form measures of intersection density and block length) and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization. Finally, it considers what these indicators suggest about the urban form and transportation infrastructure in different cities.

@@

Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide. This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents. The 50 sampled cities span Africa, Asia, Australia, Europe, and North and South America. We look separately at each city's drivable and walkable street networks. For cities such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we take the modern CBD as the city center.

@@

\begin{figure*}[htbp]
	\center
	\includegraphics[width=\textwidth]{figures_tables/world_map_bw.png}
	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by drivable $\phi$ terciles (lower values mean less planar).}
	\label{fig:world_map_bw}
\end{figure*}

\begin{figure}[htbp]
	\includegraphics[width=0.48\textwidth]{figures_tables/regression_lambda_phi_linear.png}
	\caption{Plot of $\lambda$ vs $\phi$ by network type with simple regression lines (dashed=drive, dotted=walk).}
	\label{fig:regression_lambda_phi}
\end{figure}

@@

We can see the impact of this bias in aggregate across our study sites: the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks). To summarize, planarity varies both across cities as well as across different neighborhoods within individual cities. Modeling urban street networks as planar graphs can bias urban form analyses in several ways:

@@

Future research can further explore how these indicators correlate with other measures of urbanization, development, investment, technology, politics, and era. While this study focused on measures of the urban form, particularly intersection counts and block lengths, future research can also further explore impacts of planar simplifications on topological analyses and routing.

@@

Street networks are often defined and modeled as planar graphs in the urban studies literature --- despite typically being nonplanar in reality --- to simplify real-world complexity and make certain analyses tractable. This study measured the formal and spatial planarity of central urban street networks empirically, demonstrating how planar simplifications inconsistently affect urban form analysis results. It developed two new measures of the extent to which spatial networks are nonplanar --- the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality. Although planar graphs are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and block lengths. Individual urban street network analyses must consider these location-dependent tractability and accuracy trade-offs when choosing between planar and nonplanar models.

>>>>

\usepackage[final]{draftwatermark}
\SetWatermarkText{DRAFT}
\SetWatermarkScale{1}
\SetWatermarkLightness{0.9}

@@

\runninghead{Author Redacted}
\title{Planarity and Street Network Representation in Urban Form Analysis}
\author{Author Redacted\affilnum{1}}
\affiliation{\affilnum{1}Affiliation redacted}
\corrauth{Author Redacted, Address Redacted}
\email{Email Redacted}

@@

\keywords{GIS, street networks, transportation, urban form, urban morphology}

@@

In urban planning and transportation research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity. Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,xiao_identifying_2016,zhong_revealing_2017}.

@@

\section{Background and motivation}

\subsection{Street network models}

@@

\begin{figure}[tbp]
	\center
	\includegraphics[width=0.49\textwidth]{figures_tables/planar_vs_not.png}
	\caption{Formally and spatially planar (left) and nonplanar (right) road networks.}
	\label{fig:planar_vs_not}
\end{figure}

@@

\subsection{Planar models in urban research}

In the urban studies literature, street networks are commonly modeled as planar graphs. Many street networks are obviously nonplanar in reality, but researchers use planar models as a potentially useful simplification of real-world complexity. Many scholars argue that street networks are planar graphs (e.g., \citealp[p.~18]{batty_network_2005}; \citealp[p.~521]{buhl_topological_2006}; \citealp[p.~1]{hu_topological_2008}; \citealp[p.~259]{masucci_random_2009}; \citealp[p.~114]{porta_networks_2010}; \citealp[p.~3]{strano_elementary_2012}; \citealp[p.~1]{masucci_limited_2013}; \citealp[p.~1074]{strano_urban_2013}; \citealp[p.~168]{law_defining_2017}). Others prefer to hedge slightly, arguing that street networks are approximately or essentially planar --- close enough to be well-modeled as such (e.g., \citealp[p.~6]{dill_measuring_2004}; \citealp[p.~3]{cardillo_structural_2006};  \citealp[p.~340]{xie_measuring_2007}; \citealp[p.~1]{barthelemy_modeling_2008}; \citealp[p.~3]{barthelemy_spatial_2011}; \citealp[pp.~563]{chan_urban_2011}; \citealp[p.~1]{gudmundsson_entropy_2013}; \citealp[p.~1]{viana_simplicity_2013}; \citealp[p.~2]{louf_typology_2014}; \citealp[p.~2191]{zhong_detecting_2014}; \citealp[p.~2]{wang_resilience_2015}; \citealp[p.~42]{aldous_routed_2016}; \citealp[p.~257]{barthelemy_paths_2017}).

@@

If a planar graph models a street network poorly, it could do so in multiple ways. Imposing planarity on a nonplanar street network produces artificial nodes in the graph at bridges and tunnels, which breaks routing. For example, this misrepresentation would allow a routing engine to recommend a left turn at an \enquote{intersection} that is actually an overpass \citep[p.~6]{kwan_review_1996}. Intersection counts and densities would be overestimated in the presence of these false nodes, and edge lengths would be underestimated due to these artificial breakpoints splitting up street segments. This consequently impacts interpretation of urban form density, grain, pattern, connectedness, and permeability. Finally, this bias might behave inconsistently across different kinds of cities and street network types based on the extents to which they are planar.

\subsection{Open questions}

Given these issues, some unanswered questions remain in the empirical literature. What do \emph{approximately planar} and \emph{well-modeled} mean for street network research? How do planar simplifications impact the results of street network analyses? Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities (or neighborhoods) to different extents? And if street networks are at least generally approximately planar, how can we measure the extent to which a given street network is nonplanar?

The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,chimani_non-planar_2009}. However, these measures are imperfect and hard to compute \citep{szekely_successful_2004,chimani_vertex_2012}. They also fail to adjust for the size, density, or real-world embedding of a spatial network. Discussing road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks no widely-accepted measure of the extent of nonplanarity has emerged and calls for the development of better indicators.

@@

This study builds on this rich body of street network research by developing two new measures of the extent to which a spatial network is planar. It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results (focusing on the common urban form measures of intersection density and street segment length) and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization. Finally, it considers what these indicators suggest about the urban form and transportation infrastructure in different cities.

@@

Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide. This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents \citep[cf.][]{horner_excess_2002,gil_street_2017}. The 50 sampled cities span Africa, Asia, Australia, Europe, and North and South America. We look separately at each city's drivable and walkable street networks. For cities such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we take the modern CBD as the city center.

@@

\begin{figure*}[htbp]
	\centering
	\includegraphics[width=\textwidth]{figures_tables/world_map_bw.png}
	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by drivable $\phi$ terciles (lower values mean less planar).}
	\label{fig:world_map_bw}
\end{figure*}

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.49\textwidth]{figures_tables/regression_lambda_phi_linear.png}
	\caption{Plot of $\lambda$ vs $\phi$ by network type with regression lines.}
	\label{fig:regression_lambda_phi}
\end{figure}

@@

We can see the impact of this bias in aggregate across our study sites: the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks). These impact our understanding of the urban form's density, grain, pattern, connectedness, and permeability. To summarize, planarity varies both across cities as well as across different neighborhoods within individual cities. Modeling urban street networks as planar graphs can bias urban form analyses in several ways:

@@

Future research can further explore how these indicators correlate with other measures of urbanization, development, investment, technology, politics, and era. While this study focused on measures of the urban form, particularly intersection counts and street segment lengths, future research can also further explore impacts of planar simplifications on topological analyses and routing.

@@

Street networks are often defined and modeled as planar graphs in the urban studies literature --- despite typically being nonplanar in reality --- to simplify real-world complexity and make certain analyses tractable. This study measured the formal and spatial planarity of central urban street networks empirically, demonstrating how planar simplifications inconsistently affect urban form analysis results. It developed two new measures of the extent to which spatial networks are nonplanar --- the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality. Although planar graphs are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and street lengths. Individual urban street network analyses must consider these location-dependent tractability and accuracy trade-offs when choosing between planar and nonplanar models.

\section{Acknowledgments}

Redacted.
